---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd for corrections file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
library(tidyverse)
library(lubridate)
```

```{r automated, include = FALSE}
# Automated process
source("0-data/0-L2M-download-2019-20.R") # Download the L2M reports
source("0-data/0-bkref-data.R") # Download bkref box scores
source("1-tidy/1-L2M-raw.R") # Gather up and tidy the L2M reports
source("1-tidy/1-L2M-bkref.R") # Combine tidy L2Ms with bkref box scores
```

# L2M

## Articles

- [How To Extract The Last Two Minute Reports](2-eda/2-how-to-last-two-minutes)
- [Updating The Last Two Minute Reports: data through the 2019 NBA Finals](2-eda/2-through-2019-finals)


## Description

This repository is meant as a way to maintain the Last Two Minute (L2M) reports that the NBA releases for certain NBA games. The clean, processed version of the data can be found in [1-tidy/L2M/L2M.csv](1-tidy/L2M/L2M.csv) while the corresponding `R` code to create this data is scattered across the [0-data](0-data) and [1-tidy](1-tidy) folders. The [0-data/L2M/](0-data/L2M/) folder hosts the raw data (in pdf form).

```{r calls, echo = FALSE}
l2m <- read_rds("1-tidy/L2M/L2M.rds") %>% 
  mutate(decision_2 = ifelse(is.na(decision) | decision == "",
                             "INC", decision),
         decision_3 = ifelse(is.na(decision) | decision == "",
                             "inc", decision),
         correct = ifelse(decision_2 %in% c("CC", "CNC"),
                          "correct", "incorrect"),
         correct_num = 1*(correct == "correct"),
         foul = ifelse(decision_2 %in% c("CC", "INC"),
                       "foul", "clean"),
         foul_num = 1*(foul == "foul"),
         temp_year = ifelse(month(date) < 10, 2016, 2015),
         temp_date = as.Date(paste(temp_year, month(date),
                                   day(date), sep = "/"), "%Y/%m/%d"))


games <- l2m %>% 
  group_by(temp_date, season) %>% 
  summarise(games = n_distinct(bkref_id),
            quarters = n_distinct(bkref_id, period))

l2m_freq <- l2m %>%
  group_by(temp_date, season) %>% 
  count(decision_2) %>% 
  spread(decision_2, n) %>% 
  replace(is.na(.), 0) %>% 
  mutate(calls = CC + CNC + IC + INC) %>% 
  left_join(games)

##########
l2m_freq %>% 
  ggplot(aes(temp_date, calls/quarters, color = factor(season),
             group = season)) +
  geom_smooth(aes(linetype = factor(season)), method = "loess") +
  geom_vline(xintercept = as.numeric(as.Date("2016-04-15")),
             linetype = 4, color = "black") +
  #scale_y_continuous(labels=percent) +
  scale_color_viridis_d() +
  labs(color = "Season", linetype = "Season", x = "", y = "",
       title = "Number of Graded Action",
       subtitle = "per period graded",
       caption = paste0("@atlhawksfanatic", "\nL2M Reports through ", max(l2m$date))) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r whistles, echo = FALSE}
l2m %>%
  rename(Season = season) %>% 
  group_by(Season) %>%
  summarise(Games = n_distinct(bkref_id),
            "Grades per period" = round(n() / n_distinct(bkref_id, period), digits = 2),
            "Calls per period" = round((sum(decision_3 == "CC") + sum(decision_3 == "IC")) / n_distinct(bkref_id, period), digits = 2),
            "IC per period" = round(sum(decision_3 == "IC") / n_distinct(bkref_id, period), digits = 2),
            "CC Percentage" = scales::percent(sum(decision_3 == "CC") / (sum(decision_3 == "CC") + sum(decision_3 == "IC"))),
            "INC per period" = round((sum(decision_3 == "INC") + sum(decision_3 == "inc")) / n_distinct(bkref_id, period), digits = 2),
            "Bad Calls Percentage" = scales::percent((sum(decision_3 == "IC") + sum(decision_3 == "INC") + sum(decision_3 == "inc")) / (sum(decision_3 == "CC") + sum(decision_3 == "IC") + sum(decision_3 == "INC") + sum(decision_3 == "inc"))),
            "CNC per period" = round(sum(decision_3 == "CNC") / n_distinct(bkref_id, period), digits = 2)) %>% 
  knitr::kable(caption = paste0("L2M Call Accuracy through ", max(l2m$date)), align = "c")

```

The process for compiling the L2M dataset is to:

1. Download the raw data, these are broken up into the different years which the NBA has collected L2M:
    - [Archived](0-data/0-L2M-download-archive.R) which begins on 1 March 2015 and goes through the 2017 NBA Finals.
    - [2017-18](0-data/0-L2M-download-2017-18.R)
    - [2018-19](0-data/0-L2M-download-2018-19.R) - which changes formats to online only after the 2019 NBA All Star Game (February 21, 2019 is the first). This requires the [splashr](https://github.com/hrbrmstr/splashr) package to handle scraping of the NBA website.
    - [2019-20](0-data/0-L2M-download-2019-20.R) - current season and so far no PDFs. [splashr](https://github.com/hrbrmstr/splashr) package required.
2. Read in the pdf files through the [pdftools](https://github.com/ropensci/pdftools) package:
    - [Archived](0-data/0-L2M-pdftools-archive.R)
    - [2017-18](0-data/0-L2M-pdftools-2017-18.R)
    - [2018-19](0-data/0-L2M-pdftools-2018-19.R)
    - None for 2019-20 at the moment.
3. Download box scores for games from <basketball-reference.com> for score and rosters to match up committing/disadvantaged players.
    - [File](0-data/0-bkref-data.R) depends on .rds files created in step 2 to be present in directory.
4. Combine L2M reports with box score information
    - [Raw version](1-tidy/1-L2M-raw.R) which does not include box score info, the [csv file](1-tidy/L2M/L2M_raw.csv)
    - [Final version](1-tidy/1-L2M-bkref.R) which includes box score info, the [csv file](1-tidy/L2M/L2M.csv)

The final [output](1-tidy/L2M/L2M.csv) includes the following variables:

- `period`: period at which point the play occurred
- `time`: time remaining in the period when play occurred
- `call_type`: raw call type variable in L2M
- `committing`: committing player or team in L2M, may be blank
- `disadvantaged`: disadvantaged player or team in L2M, may be blank
- `decision`: judgement of L2M for the call, could be CC, CNC, IC, INC, or blank where CC = Correct Call, CNC = Correct Non-Call, IC = Incorrect Call, INC = Incorrect Non-Call and blank = not detectable without technology
- `comments`: L2M comments on the play
- `game_details`: game details on L2M
- `page`: page of L2M for pdf
- `file`: name of L2M file, will be NA for scraped data
- `game_date`: game date according to L2M report header
- `away_team`: away team name
- `home_team`: home team name
- `call`: first part of call_type, this is before the colon in call_type
- `type `: second part of call_type, this is after the colon in call_type
- `date`: date of game in YYYY-MM-DD format
- `away`: away team abbreviation
- `home`: home team abbreviation
- `scrape_time`: time that NBA website was scraped for L2M
- `stint`: stint which indicates when a set of plays roughly occurred, only available for scraped data
- `home_bkref`: home team abbreviation according to basketball-reference
- `bkref_id`: game id for basketball-reference
- `ref_1`: name of first referee for game
- `ref_2`: name of second referee for game
- `ref_3`: name of third referee for game
- `attendance`: attendance for the game
- `committing_min`: total minutes played by player committing action (note, may be NA because the player did not play and likely an input error from NBA on L2M)
- `committing_team`: team for committing player
- `committing_side`: home/away for committing player
- `disadvantaged_min`: total minutes played by player disadvantaged by action
- `disadvantaged_team`: team for disadvantaged player
- `disadvantaged_side`: home/away for disadvantaged player
- `type2`: consistent format for type of infraction
- `time_min`: minutes remaining in period
- `time_sec`: seconds remaining in period
- `time2`: fractional minutes left (ie 1.9 would be one minute and 54 seconds)
- `season`: NBA season for which the graded play was a part of, convention is to use the last year of the NBA season so 2015 refers to the 2014-15 Season
- `playoffs`: dummy variable equal to 1 if the game occurred in the playoffs

And an overview of the changes in L2M reporting is provided in [2-eda/2-through-2019-finals](2-eda/2-through-2019-finals). Further, there is a bit of a how-to for downloading and extracting data on the L2M that is provided in [2-eda/2-how-to-last-two-minutes](2-eda/2-how-to-last-two-minutes)